The Soulgem Paper

### 2011-09-24 soulgem.tex
Second draft of expectations table. Now with a respectable caption :-P.
Second draft of Method and Study Execution section.

Another reason for role evolution arises from changes in the expectations that are needed to meet the organization's goals. An involvement in a new market niche may bring the need to coordinate the old products with the new one, to transfer expertise, and so on. One generally accepted cause of shifts in expectations is organizational growth \cite{Blau1971}. With growth comes specialization, and specialization consists of narrower role definitions. Therefore, other things being equal, we expect larger organizations to have a greater degree of role specialization.

@@

We performed a qualitative empirical study based on the theoretical framework presented above to improve our understanding of roles in software organizations. Since there is already a significant amount of theoretical work on the concept of roles in organizations, we did not need a grounded theory approach. We therefore chose to follow Yin's case study methodology \cite{Yin2003}. We executed an explanatory two-case case study with two embedded units of analysis (software organizations and roles).

@@

\textbf{\emph{RQ3:}} Why are expectations clustered in roles in the way they are? That is, what is the logic, if any, behind organizing expectations in the particular ways in which they have been, and not in others?

\emph{Proposition:} Expectations are clustered in roles based on a combination of institutional, historical, structural, and personal factors. Institutional factors refer to the larger culture: what we generally expect a person fulfilling a certain role should be doing. Historical factors consist of past experiences in the organization: what people fulfilling this role or similar roles have typically done in the past. Structural factors refer to the power and affordances available to a person by virtue of their position in the social structure. Personal factors include the preferences and background of the agents enacting the role.

\textbf{\emph{RQ4:}} Is there a significant divergence between roles and positions in software organizations? If so, why, and what are its consequences?

\emph{Proposition:} There is a significant divergence between roles and positions, and the reason is that position is a structural factor, as specified above, and there are many other factors that contribute to the shaping of a role.

@@

\emph{Proposition:} There are two main drivers of role evolution: organizational growth, which leads to specialization \cite{Blau1971,Haveman1993}, and organizational dysfunction, which leads to rearrangements of expectations.

@@

To answer our research questions, we collected field data from two software organizations in the area of Victoria, Canada. Our case selection was opportunistic: we wanted groups that develop software as their main activity, and whose headquarters were located in our vicinity. We contacted managers in both organizations and negotiated access to their sites in return for feedback on their development practices.

We carried out our case study during the Winter and Spring of 2011. The first author spent about six weeks performing observations, three weeks at each of the two sites, sitting at a desk in their development areas, and attending to stand-up meetings, product meetings, all-hands meetings, and several other gatherings and social events. He also gained user access to the issue tracking and project estimation systems at both sites, and manually examined several issues, bugs, and milestones.

@@

Following Yin's \cite{Yin2003} suggestions, we analyzed all the observation and interview data examining whether it supported or challenged our study propositions. Whenever possible we triangulated our evidence. We were also particularly interested in evidence that did not fit with our theoretical framework, in order to refine the latter for future studies.

>>>>

Another reason for role evolution arises from changes in the expectations that are needed to meet the organization's goals. An involvement in a new market niche may bring the need to coordinate the old products with the new one, to transfer expertise, and so on. One generally accepted cause of shifts in expectations is organizational growth \cite{Blau1971,Haveman1993}. With growth comes specialization, and specialization consists of narrower role definitions. Therefore, other things being equal, we expect larger organizations to have a greater degree of role specialization.

@@

We performed a qualitative empirical study based on the theoretical framework presented above to improve our understanding of roles in software organizations. We chose to follow Yin's case study methodology \cite{Yin2003}. We executed an explanatory two-case case study with two embedded units of analysis (software organizations and roles).

@@

\textbf{\emph{RQ3:}} What is the logic (what are the factors) behind the particular clusters of expectations found in software organizations?

\emph{Proposition:} Expectations are clustered by institutional and organizational factors, but we do not know the specific reasons that drive this clustering.

\textbf{\emph{RQ4:}} Is there a significant divergence between roles and positions in software organizations? Why?

\emph{Proposition:} There is a significant divergence between roles and positions, and the reason is that position is an institutional factor, as specified above, and there are many other factors that contribute to the shaping of a role.

@@

\emph{Proposition:} There are two main drivers of role evolution: organizational growth, which leads to specialization, and organizational dysfunction, which leads to rearrangements of expectations.

@@

We collected field data from two software organizations in the area of Victoria, Canada. Our case selection was opportunistic: we wanted groups that develop software as their main activity, and whose headquarters were located in our vicinity. We contacted managers in both organizations and negotiated access to their sites in return for feedback on their development practices.

We carried out our case study during the Winter and Spring of 2011. The first author spent five weeks performing observations, three weeks at the first and two at the second site, sitting at a desk in their development areas, and attending to several kinds of meetings (stand-up, product, all-hands, and others). He also gained access to the issue tracking and project estimation systems at both sites, and manually examined several issues, bugs, and milestones.

@@

Following Yin's \cite{Yin2003} suggestions, we analyzed all the observation and interview data examining whether it supported or challenged our study propositions. Whenever possible we triangulated our evidence. We were also particularly interested in evidence that did not fit with our theoretical framework, in order to refine it for future studies.
